Stochastic Variance Reduced Policy Gradient

### 2018-01-11 svrpg.tex
added svrpg pseudocode

\begin{algorithm}[tb]
	\caption{SVRG}
	\label{alg:svrg}
	\begin{algorithmic}
		\STATE {\bfseries Input:} number of epochs $S$, epoch size $m$, step size $\alpha$, initial parameter $\vtheta_0$
		\FOR{$s=0$ {\bfseries to} $S-1$}
		\STATE $\vtheta_{s+1}^0 = \tilde{\vtheta} = \vtheta_s$
		\FOR{$t=0$ {\bfseries to} $m-1$}
		\STATE Pick $i$ uniformly at random from $[1,N]$
		\STATE $\blacktriangledown f(\vtheta_{s+1}^t) = 
					\nabla f(\tilde{\vtheta}) + 
					\nabla g(x_i\mid\vtheta_{s+1}^t) -
					\nabla g(x_i \mid \tilde{\vtheta})$
		\STATE $\vtheta_{s+1}^{t+1} = \vtheta_{s+1}^{t} + \alpha\blacktriangledown f(\vtheta_{s+1}^t)$
		\ENDFOR
		\STATE $\vtheta_{s+1} = \vtheta_{s+1}^m$
		\ENDFOR
		\STATE {\bfseries Convex case: return} $\vtheta_S$
		\STATE {\bfseries Non-Convex case: return} $\vtheta_s^t$ with $(s,t)$ picked uniformly at random from $[0,S)\times[0,m)$
	\end{algorithmic}
\end{algorithm}

\section{Algorithm}
In on-line policy gradient problems, stochastic gradient ascent (SGA) is not really a choice, but is dictated by the necessity of interacting with an unknown environment. We would like to apply SVRG to the policy gradient framework in order to limit the variance introduced by sampling trajectories, which would ultimately lead to faster convergence. In this scenario, the function $f(\vtheta)$ to optimize is $J(\vtheta)$ and the (implicit) dataset $\mathcal{D}$ is the set of all possible trajectories $\mathcal{T}$. Compared to the typical finite sum optimization scenario, we have the following additional challenges:
\begin{itemize}
	\item The objective $J(\vtheta)$ is typically non-convex;
	\item The dataset $\mathcal{T}$ can be infinite;
	\item The value of the parameter $\vtheta$ influences the sampling of trajectories;
\end{itemize}
To deal with non-convexity, we require $J(\vtheta)$ to be L-smooth, which is a reasonable assumption for common policy classes such as truncated Gaussian and Softmax [appendix?]. Because of the infinite dataset, we can only rely on an estimate of the full gradient. Finally, we employ importance weighting \cite{rubinstein1981simulation} \cite{precup2000eligibility} to guarantee proper sampling of trajectories. We focus on the REINFORCE gradient estimator for simplicity, but the extension to G(PO)MDP is straightforward. The proposed SVRG-like gradient estimate, which we call Stochastic Variance Reduced Policy Gradient (SVRPG), is the following:
\begin{align*}
	\blacktriangledown J(\vtheta) = &\frac{1}{N}\sum_{j=0}^{N-1}\nabla\log\pi(\tau_j \mid \tilde{\vtheta}) \\
		&+ \omega(\tau_i)\nabla\log\pi(\tau_i \mid \vtheta) 
		- \nabla\log\pi(\tau_i \mid \tilde{\vtheta}),
\end{align*}
where trajectories $\tau_j$ for $j=1,\dots,N$ and $\tau_i$ are all sampled independently from the snapshot policy $\pi_{\tilde{\vtheta}}$, and $\omega(\tau) = \frac{\pi(\tau\mid\vtheta)}{\pi(\tau\mid\tilde{\vtheta})}$ is an importance weight from $\pi_{\tilde{\vtheta}}$ to the current policy $\pol$. We immediately have the following (proof in the Appendix):
\begin{restatable}{lemma}{unbias}\label{lemma:unbias}
The SVRPG gradient estimator is unbiased:
\[
\mathop{E}_{\tau\sim\pi(\cdot\mid\tilde{\vtheta})}\left[\blacktriangledown J(\vtheta)\right] = \gradJ{}.
\]
\end{restatable}
The resulting SVRPG algorithm is detailed in [todo:pseudocode].
[Next: explain which details are left open and briefly outline next sections; deeper explanation of algorithm?]

>>>>

\begin{algorithm}[tb]
	\caption{SVRG}
	\label{alg:svrg}
	\begin{algorithmic}
		\STATE {\bfseries Input:} number of epochs $S$, epoch size $m$, step size $\alpha$, initial parameter $\vtheta_{m-1}^0$
		\FOR{$s=0$ {\bfseries to} $S-1$}
		\STATE $\vtheta_0^{s+1} = \tilde{\vtheta}^s = \vtheta_{m-1}^s$
		\FOR{$t=0$ {\bfseries to} $m-1$}
		\STATE Pick $i$ uniformly at random from $[1,N]$
		\STATE \begin{align*}
			\blacktriangledown f(\vtheta_t^{s+1}) = 
			&\nabla f(\tilde{\vtheta}^s) \\ + 
			&\nabla g(x_i\mid\vtheta_t^{s+1}) -
			\nabla g(x_i \mid \tilde{\vtheta}^{s+1})
		\end{align*}
		\STATE $\vtheta_{t+1}^{s+1} = \vtheta_t^{s+1} + \alpha\blacktriangledown f(\vtheta_t^{s+1})$
		\ENDFOR
		\ENDFOR
		\STATE {\bfseries Convex case: return} $\vtheta_{m-1}^S$
		\STATE {\bfseries Non-Convex case: return} $\vtheta_t^s$ with $(s,t)$ picked uniformly at random from $[0,S)\times[0,m)$
	\end{algorithmic}
\end{algorithm}

\section{Algorithm}
In on-line policy gradient problems, stochastic gradient ascent (SGA) is not really a choice, but is dictated by the necessity of interacting with an unknown environment. We would like to apply SVRG to the policy gradient framework in order to limit the variance introduced by sampling trajectories, which would ultimately lead to faster convergence. In this scenario, the function $f(\vtheta)$ to optimize is $J(\vtheta)$ and the (implicit) dataset $\mathcal{D}$ is the set of all possible trajectories $\mathcal{T}$. Compared to the typical finite sum optimization scenario, we have the following additional challenges:
\begin{itemize}
	\item The objective $J(\vtheta)$ is typically non-convex;
	\item The dataset $\mathcal{T}$ can be infinite;
	\item The value of the parameter $\vtheta$ influences the sampling of trajectories;
\end{itemize}
To deal with non-convexity, we require $J(\vtheta)$ to be L-smooth, which is a reasonable assumption for common policy classes such as truncated Gaussian and Softmax [appendix?]. Because of the infinite dataset, we can only rely on an estimate of the full gradient. Finally, we employ importance weighting \cite{rubinstein1981simulation} \cite{precup2000eligibility} to guarantee proper sampling of trajectories. We focus on the REINFORCE gradient estimator for simplicity, but the extension to G(PO)MDP is straightforward. The proposed SVRG-like gradient estimate, which we call Stochastic Variance Reduced Policy Gradient (SVRPG), is the following:
\begin{align*}
	\blacktriangledown J(\vtheta) = &\frac{1}{N}\sum_{j=0}^{N-1}\nabla\log\pi(\tau_j \mid \tilde{\vtheta}) \\
		&+ \omega(\tau_i)\nabla\log\pi(\tau_i \mid \vtheta) 
		- \nabla\log\pi(\tau_i \mid \tilde{\vtheta}),
\end{align*}
where trajectories $\tau_j$ for $j=1,\dots,N$ and $\tau_i$ are all sampled independently from the snapshot policy $\pi_{\tilde{\vtheta}}$, and $\omega(\tau) = \frac{\pi(\tau\mid\vtheta)}{\pi(\tau\mid\tilde{\vtheta})}$ is an importance weight from $\pi_{\tilde{\vtheta}}$ to the current policy $\pol$. We immediately have the following (proof in the Appendix):
\begin{restatable}{lemma}{unbias}\label{lemma:unbias}
The SVRPG gradient estimator is unbiased:
\[
\mathop{E}_{\tau\sim\pi(\cdot\mid\tilde{\vtheta})}\left[\blacktriangledown J(\vtheta)\right] = \gradJ{}.
\]
\end{restatable}
The resulting SVRPG algorithm is detailed in Algorithm \ref{alg:svrpg}.
[Next: explain which details are left open and briefly outline next sections; deeper explanation of algorithm?]

\begin{algorithm}[tb]
	\caption{SVRPG}
	\label{alg:svrpg}
	\begin{algorithmic}
		\STATE {\bfseries Input:} number of epochs $S$, epoch size $m$, step size $\alpha$, initial parameter $\vtheta_{m-1}^0$, batch size $N$
		\FOR{$s=0$ {\bfseries to} $S-1$}
		\STATE Sample N trajectories $\tau_j$ from $\pi(\cdot\mid\tilde{\vtheta}^{s})$
		\STATE $\gradApp{\tilde{\vtheta}^{s}}{N} = \frac{1}{N}\sum_{j=0}^{N-1}\score{}{\tau_j\mid\tilde{\vtheta}^{s}}$
		\STATE $\vtheta_0^{s+1} = \tilde{\vtheta}^s = \vtheta_{m-1}^s$
		\FOR{$t=0$ {\bfseries to} $m-1$}
		\STATE Sample trajectory $\tau_i$ from 				$\pi(\cdot\mid\tilde{\vtheta}^{s})$
		\STATE 
		\begin{align*}
		\blacktriangledown J(\vtheta_t^{s+1}) = 
		&\gradApp{\tilde{\vtheta}^s}{N}
		+ \omega(\tau_i)\score{}{t_i\mid\vtheta_t^{s+1}} \\
		& - \score{}{\tau_i \mid \tilde{\vtheta}^{s+1}}
		\end{align*}
		\STATE $\vtheta_{t+1}^{s+1} = \vtheta_t^{s+1} + \alpha\blacktriangledown J(\vtheta_t^{s+1})$
		\ENDFOR
		\ENDFOR
		\STATE {\bfseries return} $\vtheta_t^s$ with $(s,t)$ picked uniformly at random from $[0,S)\times[0,m)$
	\end{algorithmic}
\end{algorithm}
